Why is it that statistical tests for residuals are not widely used? An explanation using visual inference.

### 2023-02-08 paper_comparison/paper_comparison/paper_comparison.Rmd
draft of conclusion more complete

1. summarise paper content

This paper has demonstrated that conventional residual-based statistical tests are more sensitive to departures from assumptions for residuals than visual tests evaluated by humans in the diagnostics of linear regression models. An experiment is conducted in three data collection periods using two simulation models, one for for non-linearity and and the other for heteroskedasticity. Methods for obtaining the $p$-value of visual tests proposed by @vanderplas2021statistical are extended to allow for multiple selections in a lineup protocol.

2. briefly describe what was learned

According to the data collected from the human subject experiment, residual plots rejected by visual tests are rarely not rejected by conventional tests, but departures of residuals that are of little practical sense to humans will often be alarmed by conventional tests, supporting the common guidance given by experts to utilize graphical representations of data in model diagnostics.

We have also found that residual-based conventional tests relying on a user-parameterized test formula, such as the RESET test, will suffer from a significant decrease in power for detecting certain residual departures intended to be tested but not covered by the test formula. This can not be avoided without prior knowledge of their functional forms, which is usually obtained by plotting the residuals. On the other hand, visual tests do not rely on such information and a collection of departures can be tested simultaneously.

@@

orientation of heter shape

>>>>

<!--1. summarise paper content-->

Motivated by the advice of regression analysis experts, that residual plots as opposed to conventional tests are an indispensible methods for assessing model fit, a human subjects experiment was conducted using visual inference. The experiment tested two primary departures from good residuals: non-linearity and heteroskedasticity.

<!-- This paper has demonstrated that conventional residual-based statistical tests are more sensitive to departures from assumptions for residuals than visual tests evaluated by humans in the diagnostics of linear regression models. An experiment is conducted in three data collection periods using two simulation models, one for for non-linearity and and the other for heteroskedasticity. Methods for obtaining the $p$-value of visual tests proposed by @vanderplas2021statistical are extended to allow for multiple selections in a lineup protocol.-->

<!--2. briefly describe what was learned-->

The experiment found that conventional residual-based statistical tests are more sensitive to departures from assumptions for residuals than visual tests as would be evaluated by humans. That is, conventional tests conclude there are problems with the model fit almost twice as often as humans would. They often reject when departures in the form of non-linearity and heteroskedasticity are not visible to a human.

One might say that this is correct behaviour, but it can be argued that the conventional tests are rejecting when it is not necessary. Many of these rejections happen even when downstream analysis and results would not be significantly affected by the small departures from a good fit. The results from human evaluations provide a more practical solution, which reinforces the statements from regression experts that residual plots are an indispensible method for model diagnostics.

Now it is important to note that residual plots need to be delivered as a line up, where it is embedded in a field of null plots. This enables a careful calibration for reading structure in residual plots. XXX More here

However, human evaluation of residuals is expensive. It is time-consuming, and laborious. This is another reason why it often appears to be ignored. With the availability of sophisticated computer vision algorithms today, the goal of this work is to form the basis of providing automated residual plot reading. XXX More here

The experiment also revealed some interesting details about how residual plots are read. For the most part, the visual test performed very similarly to the (best) conventional test only with the power curve shifted in the less sensitive direction. Unlike the conventional tests, where one needs to specifically test for non-linearity or heteroskedasticity the visual test operated effectively across the range of departures from good residuals.

As expected, if the fitted value distribution is not uniform, there is a loss of power in the visual test. Structure is hardest to detect if fitted values ar discrete. Also, a surprising find was that the direction of heteroskedasticity appears to affect the ability to visually detect it, with wedge to the left being less detectable.

<!--According to the data collected from the human subject experiment, residual plots rejected by visual tests are rarely not rejected by conventional tests, but departures of residuals that are of little practical sense to humans will often be alarmed by conventional tests, supporting the common guidance given by experts to utilize graphical representations of data in model diagnostics. -->

<!-- We have also found that residual-based conventional tests relying on a user-parameterized test formula, such as the RESET test, will suffer from a significant decrease in power for detecting certain residual departures intended to be tested but not covered by the test formula. This can not be avoided without prior knowledge of their functional forms, which is usually obtained by plotting the residuals. On the other hand, visual tests do not rely on such information and a collection of departures can be tested simultaneously.

@@

orientation of heter shape
-->

### 2023-02-08 paper_comparison/paper_comparison/paper_comparison.tex
draft of conclusion more complete

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  summarise paper content
\end{enumerate}

This paper has demonstrated that conventional residual-based statistical
tests are more sensitive to departures from assumptions for residuals
than visual tests evaluated by humans in the diagnostics of linear
regression models. An experiment is conducted in three data collection
periods using two simulation models, one for for non-linearity and and
the other for heteroskedasticity. Methods for obtaining the \(p\)-value
of visual tests proposed by \citet{vanderplas2021statistical} are
extended to allow for multiple selections in a lineup protocol.

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\setcounter{enumi}{1}
\tightlist
\item
  briefly describe what was learned
\end{enumerate}

According to the data collected from the human subject experiment,
residual plots rejected by visual tests are rarely not rejected by
conventional tests, but departures of residuals that are of little
practical sense to humans will often be alarmed by conventional tests,
supporting the common guidance given by experts to utilize graphical
representations of data in model diagnostics.

We have also found that residual-based conventional tests relying on a
user-parameterized test formula, such as the RESET test, will suffer
from a significant decrease in power for detecting certain residual
departures intended to be tested but not covered by the test formula.
This can not be avoided without prior knowledge of their functional
forms, which is usually obtained by plotting the residuals. On the other
hand, visual tests do not rely on such information and a collection of
departures can be tested simultaneously.

Interestingly, the orientation of the visual pattern

fitted value distribution

orientation of heter shape

>>>>

Motivated by the advice of regression analysis experts, that residual
plots as opposed to conventional tests are an indispensible methods for
assessing model fit, a human subjects experiment was conducted using
visual inference. The experiment tested two primary departures from good
residuals: non-linearity and heteroskedasticity.

The experiment found that conventional residual-based statistical tests
are more sensitive to departures from assumptions for residuals than
visual tests as would be evaluated by humans. That is, conventional
tests conclude there are problems with the model fit almost twice as
often as humans would. They often reject when departures in the form of
non-linearity and heteroskedasticity are not visible to a human.

One might say that this is correct behaviour, but it can be argued that
the conventional tests are rejecting when it is not necessary. Many of
these rejections happen even when downstream analysis and results would
not be significantly affected by the small departures from a good fit.
The results from human evaluations provide a more practical solution,
which reinforces the statements from regression experts that residual
plots are an indispensible method for model diagnostics.

Now it is important to note that residual plots need to be delivered as
a line up, where it is embedded in a field of null plots. This enables a
careful calibration for reading structure in residual plots. XXX More
here

However, human evaluation of residuals is expensive. It is
time-consuming, and laborious. This is another reason why it often
appears to be ignored. With the availability of sophisticated computer
vision algorithms today, the goal of this work is to form the basis of
providing automated residual plot reading. XXX More here

The experiment also revealed some interesting details about how residual
plots are read. For the most part, the visual test performed very
similarly to the (best) conventional test only with the power curve
shifted in the less sensitive direction. Unlike the conventional tests,
where one needs to specifically test for non-linearity or
heteroskedasticity the visual test operated effectively across the range
of departures from good residuals.

As expected, if the fitted value distribution is not uniform, there is a
loss of power in the visual test. Structure is hardest to detect if
fitted values ar discrete. Also, a surprising find was that the
direction of heteroskedasticity appears to affect the ability to
visually detect it, with wedge to the left being less detectable.
